Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-10-10 article/article.tex
Commit 2024-10-10 13:12
Commit 2024-10-10 18:14

On multiple sites, cloud enhancements have been reported that exceed
momentarily the solar constant, resulting in clearness indices above
unit. A summary of extreme cloud enhancement (ECE) cases has been
compiled by \citet{Martins2022}. Cloud enhancements can have also some
practical implications. The intensity and duration of enhancements can
affect the efficiency and stability of photovoltaic power production
\citep{Lappalainen2020, Jaervelae2020}, while ECEs have the potential to
compromise the integrity of photovoltaic plants infrastructure
\citep{DoNascimento2019}. It has also been demonstrated that these
events can interfere in the comparison of ground-based and satellite
observations of radiation \citep{Damiani2018}. Global warming has likely
affected cloud coverage in the last few decades. \citet{Liu2023}
reported increases in cloud cover over the tropical and subtropical
oceans and decreases over most continents, while \citet{Dong2023}
reported decreases over North America and Europe. To our knowledge,
there is no evidence of whether this trend has also affected the number
and strength of CE events.

@@

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud-free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements have been available since 2003
from a Cimel sun photometer, which is part of the Aerosol Robotic
Network (AERONET) \citep{Giles2019, Buis1998}. From the observations in
the period 2003 -- 2023 we calculated the monthly climatological means
and standard deviations (\(\sigma\)) for the aerosol optical depth (AOD)
at \(500\,\text{nm}\) and the equivalent height of the water column
(WC). The monthly climatological values of AOD and WC, as well as
combinations with additional offsets of \(\pm1\sigma\) and
\(\pm2\sigma\), were used as inputs to the RTM in the construction of
the LUT.

@@

First, we evaluated the performance of the modeled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with
lower-than-average load of aerosols and humidity, which are the main
factors that attenuate the GHI, in the absence of clouds. With this
approach, the simulated \(\text{GHI}_\text{ref}\) should generally be
greater than the measured GHI when aerosols are more abundant. The
correlation of the \(\text{GHI}_\text{ref}\) with the GHI was tested for
a subset of 513 cloud-free days with more than \(80\,\%\) data
availability (Figure\nobreakspace{}\ref{fig:validation-GHI}). The
scatter plot shows a good agreement between the modeled and the measured
GHI with a spread that arises mainly from the short-term variability of
AOD and WC. The linear regression reveals a positive bias of
\(18.4\,\text{W}/\text{m}^2\) and a slope very close to unity. To
achieve a robust and clear distinction of the CE cases, we applied to
the \(\text{GHI}_\text{ref}\) an additional offset of
\(25\,\text{W}/\text{m}^2\).

>>>>

On multiple sites, cloud enhancements have been reported that exceed
momentarily the solar constant, resulting in clearness indices above
unit. A summary of extreme cloud enhancement (ECE) cases has been
compiled by \citet{Martins2022}. Cloud enhancements can also have some
practical implications. The intensity and duration of enhancements can
affect the efficiency and stability of photovoltaic power production
\citep{Lappalainen2020, Jaervelae2020}, while ECEs have the potential to
compromise the integrity of photovoltaic plants infrastructure
\citep{DoNascimento2019}. It has also been demonstrated that these
events can interfere in the comparison of ground-based and satellite
observations of radiation \citep{Damiani2018}. Global warming has likely
affected cloud coverage in the last few decades. \citet{Liu2023}
reported increases in cloud cover over the tropical and subtropical
oceans and decreases over most continents, while \citet{Dong2023}
reported decreases over North America and Europe. To our knowledge,
there is no evidence of whether this trend has also affected the number
and strength of CE events.

@@

Main factors responsible for the attenuation of the broadband downward
solar radiation under cloud-free atmospheres are aerosols and water
vapor. At Thessaloniki, such measurements have been available since 2003
from a Cimel sun photometer, which is part of the Aerosol Robotic
Network (AERONET) \citep{Giles2019, Buis1998}. From the observations in
the period 2003 -- 2023 we calculated the monthly climatological means
and standard deviations (\(\sigma\)) for the aerosol optical depth (AOD)
at \(500\,\text{nm}\) and the equivalent height of the water column
(WC). The monthly climatological values of AOD and WC, as well as
combinations with additional offsets of \(\pm1\sigma\) and
\(\pm2\sigma\), were used as inputs to the RTM for the construction of
the LUT. Of note, the total ozone column plays a minor role in GHI as it
affects mainly the UV wavelengths. However, it is entered in the RTM
simulations via the seasonal atmospheric profiles. The long-term
variability of total ozone in Thessaloniki for the period of study is
very small thus cannot affect the long-term \(\text{GHI}_\text{ref}\).

@@

First, we evaluated the performance of the modeled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with
lower-than-average load of aerosols and humidity, which are the main
factors that attenuate the GHI, in the absence of clouds. With this
approach, the simulated \(\text{GHI}_\text{ref}\) should generally be
greater than the measured GHI when aerosols are more abundant. The
correlation of the \(\text{GHI}_\text{ref}\) with the GHI was tested for
a subset of 513 cloud-free days with more than \(80\,\%\) data
availability (Figure\nobreakspace{}\ref{fig:validation-GHI}). The
scatter plot shows a good agreement between the modeled and the measured
GHI with a spread that arises mainly from the short-term variability of
AOD and WC. The linear regression reveals a positive bias of
\(18.4\,\text{W}/\text{m}^2\) and a slope very close to unity. To
achieve a robust and clear distinction of the CE cases, we applied to
the \(\text{GHI}_\text{ref}\) an additional offset of
\(`r C4_GLB_diff_THRES`\,\text{W}/\text{m}^2\). The long-term difference
between the measured GHI and the \(\text{GHI}_{\text{ref}}\) for the
same cloud-free days has a trend of \(0.2\,\text{W}/\text{m}^2\) per
year which further supports the validity of the produced long-term
\(\text{GHI}_\text{ref}\).
